Data Descriptor: Corresponding author journal publishing activities 2014 - 2018

### 2019-10-08 analysis/paper.Rmd
some fixes

---
title: "Data Descriptive: Corresponding author journal publishing activities 2014 - 2018"
output: github_document
bibliography: pubs.bib
csl: chicago.csl
---

@@

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, @haustein_2015 shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses [@Larivi_re_2015]. @Palzenberger_2015 provided a dataset presenting the number and proportion of scholarly articles by country of affiliation including a breakdown by author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to an open access business model is is financially feasible in terms of the global amount spent on subscriptions [@Schimmer_2015]. Since then, analyzing corresponding author publications has become a critical component of open access negotiations between library consortia and publishers.

@@

The data cannot only serve as a critical input for negotiations in the context of transformative agreements, but also in the quantitative study of the academic publishing industry and the open access uptake across countries. For instance, national transformative agreements with large publishers will increase opportunities to publish open access. The existence of such agreements will therefore likely have an impact on the open access share of a country with such agreements. To demonstrate the dataset's potential significance, we will present two new findings using the dataset: a global map of publishers's market share, and an analysis of country of affiliation by publisher. Source code used for compiling the dataset is shared along with the resulting datasets.

## Data and Methods

@@

The restriction to the article types original article and reviews reflects that in most open access license negotiations other types of journal content like letters or meeting abstracts are not considered. To allocate open access costs to institutions, @Schimmer_2015 proposed to use affiliations from corresponding authors, presuming that corresponding author affiliations can be counted straight. Under this counting method, only one institution or one country receives full credit for a publication [@Huang_2011]. However, there have been some concerns with using corresponding authorship’s data from bibliometric databases for cost analysis recently. @Gumpenberger_2018 noted two potential issues: multiple corresponding author affiliations and multiple corresponding authors per article. Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently (see Figure 1). We also found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018. Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

The underlying source code in R and SQL for both testing the retrieval and compiling the datasets are openly shared as executable reports, making our data work more reproducible and transparent. The data analytics work, which was organized as research compendium [@Marwick_2017] using the holepunch R package [@holepunch], is accessible at <https://github.com/subugoe/oa2020cadata>

@@

Overall, our dataset consists of two separate tables, which are openly available at (Link to Zenodo) as comma-separated values (csv) files and Excel spreadsheets.

@@

The file named `rp_14_18` provides a breakdown of journal information aggregated by  countries of affiliation. Only corresponding authorships, indicated by the role "reprint author" in the Web of Science database, were taken into consideration. The following table  describes the variables in detail. Overall, 99 % of records representing original articles and reviews in the period 2014 - 2018 had affiliation information for corresponding authors at the country-level.

@@

Our data confirms that the global publishing landscape is dominated by a few large publishers. In total, ten publishers accounted for around two-thirds of articles published between 2014 - 2018. The Big 3, Elsevier, Springer Nature and Wiley, accounted for a percentage of 46 %. However, further analysis suggests large differences when differentiating between subscription and fully open access journals: The market for fully open access journals seems to be far less dominated by major publishing houses. A notable exception is Springer Nature with its large fully open access portfolio including BioMed Central and the large mega-journals Scientific Reports and Nature Communications.

@@

Although the Web of Science database is a well-established bibliometric database in the quantitative science studies it comes with certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion, likely not covering the whole journal portfolio of the investigated publishers. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion.

>>>>

---
title: "Data Descriptor: Corresponding author journal publishing activities 2014 - 2018"
output: github_document
author: "Najko Jahn"
bibliography: pubs.bib
csl: chicago.csl
---

@@

Only a few openly available datasets focusing on the scholarly publishing industry exist. For example, @haustein_2015 shared an aggregated dataset that contains the market share of publishers in terms of articles indexed in the Web of Science between 1973-2013. Using this data, the authors were able to reveal a growing market concentration among a few publishing houses [@Larivi_re_2015]. @Palzenberger_2015 provided a dataset presenting the number and proportion of scholarly articles by country of affiliation including a breakdown by author role for the period 2004-2014. Published in 2015, this dataset, which was obtained from the Web of Science in-house database from the German Competence Center for Bibliometrics, formed the empirical basis for demonstrating that a large-scale transition from a subscription-based to an open access business model is is financially feasible in terms of the global amount spent on subscriptions [@Schimmer_2015]. Since then, analyzing corresponding author publications has become a critical component of open access negotiations between library consortia and publishers [@Geschuhn_2017].

@@

The data cannot only serve as a critical input for negotiations in the context of transformative agreements, but also in the quantitative study of the scholarly publishing industry and the open access uptake across countries. For instance, national transformative agreements with large publishers will increase opportunities to publish open access. The existence of such agreements will therefore likely have an impact on the open access share of a country with such agreements. To demonstrate the dataset's potential significance, we will present two new findings using the dataset: a global map of publishers's market share, and an analysis of country of affiliation by publisher. Source code used for compiling the dataset is shared along with the resulting datasets.

## Methods

@@

Like @Palzenberger_2015The, above-listed selection reflect data needs in the context of open access negotiations between library consortia and publishers. The restriction to the article types original article and reviews is in accordance with most open access contracts where other types of journal content like letters or meeting abstracts are not considered (see for instance @Sander_2019).

To allocate open access costs to institutions, @Schimmer_2015 proposed to use affiliations from corresponding authors, presuming that corresponding author affiliations can be counted straight. Under this counting method, only one institution or one country receives full credit for a publication [@Huang_2011]. However, there have been some concerns with using corresponding authorship’s data from bibliometric databases for cost analysis recently. @Gumpenberger_2018 noted two potential issues: multiple corresponding author affiliations and multiple corresponding authors per article.

Indeed, the Web of Science has begun to systematically keep track of more than one corresponding author recently: We found that the Web of Science did record more than one corresponding author for 6.2% of indexed articles published between 2014 - 2018 (see Figure 1). Moreover, a 5% of corresponding authors author was internationally co-located (see Figure 2). We therefore decided to use whole counting where every original article or review was counted once per country of affiliation of the corresponding author(s).

@@

The underlying source code in R and SQL for both testing the retrieval and compiling the datasets are openly shared as executable reports, making our data work more reproducible and transparent. The data analytics work, which was organized as research compendium [@Marwick_2017] using the holepunch R package [@holepunch], is accessible at <https://github.com/subugoe/oa2020cadata>.

More specifically, the following database retrieval steps were tracked using R Markdown, a reproducible authoring framework supporting various programming languages including R and SQL, and are accessible in the `analysis` folder of the research compendium:

1. Initial exploration about how corresponding authorships are represented in the WoS-KB ([`001_kb_rp_pretest.Rmd`](analysis/001_kb_rp_pretest.Rmd)) 
2. Coverage analysis of reprint authors in the WoS-KB for the period 2014 - 2018 ([`002_kb_rp_coverage.Rmd`](analysis/002_kb_rp_coverage.Rmd)). 
3. Steps to obtain and compile the dataset about corresponding author country affiliations 2014 - 2018([`003_kb_fetch_ca.Rmd`](analysis/003_kb_fetch_ca.Rmd)) 
4. Steps to obtain and compile the dataset about the global publisher output 2014 - 2018 ([`004_kb_fetch_publisher.Rmd`](analysis/004_kb_fetch_publisher.Rmd)).

The research compendium contains the rendered reports as well. Furthermore, this source code repository includes comprehensive reproducibility guidance for users with access to the in-house data infrastructure provided by the Competence Centre for Bibliometrics for German Science.

@@

Overall, our dataset consists of the following two separate tables:

-- `journal_publisher_14_18`: Global journal and publisher data 2014-18
-- `rp_jn_14_18`: Corresponding author country affiliations per publisher and journal 2014-18

They are provided as comma-separated values (csv) files and Excel spreadsheets.

The data are openly available in the `data/` of the research compendium hosted on GitHub <https://github.com/subugoe/oa2020cadata>.

@@

The file named `rp_jn_14_18` provides a breakdown of journal information aggregated by  countries of affiliation. Only corresponding authorships, indicated by the role "reprint author" in the Web of Science database, were taken into consideration. The following table  describes the variables in detail. Overall, 99 % of records representing original articles and reviews in the period 2014 - 2018 had affiliation information for corresponding authors at the country-level.

@@

Our data confirms that the global publishing landscape is dominated by a few large publishers. In total, ten publishers accounted for around two-thirds of articles published between 2014 - 2018. The Big 3, Elsevier, Springer Nature and Wiley, secured a market share of 46 %. However, further analysis suggests large differences between subscription and fully open access journals: The market for fully open access journals seems to be less dominated by major publishing houses. A notable exception is Springer Nature with its large fully open access portfolio including BioMed Central and the large mega-journals Scientific Reports and Nature Communications.

@@

Although the Web of Science database is a well-established bibliometric database it has certain limitations that need careful consideration when using and interpreting our data. First, the Web of Science is selective in terms of journal inclusion, likely not covering the whole journal portfolio of the investigated publishers. Although publisher names were disambiguated, journal title transfers between publishers were not tracked. Furthermore, the year of inclusion in a journal issue, i.e. print publication date, was used. However, there can be a considerable time-lag between online-first publication and issue inclusion.
